---
title: "eQTL Pipeline Report"
output: html_document
params:
  eqtl_results_filtered: NULL
  eqtl_results: NULL
  outdir: NULL
  gds_file: NULL
  single_cell_file: NULL
  counts_assay: "RNA"
  counts_slot: "counts"
  celltype_column: "CellType"
  individual_column: "Individual_ID"
  min_cells: 10
  min_expression: 0.05
  cis_distance: 1e6
  fdr_threshold: 0.05
  optimize_pcs: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(dplyr)
library(ggplot2)
library(tidyr)
```

## eQTL Analysis Report

### Pipeline Settings

The following settings were used in the pipeline:

- **Output Directory**: `r params$outdir`
- **GDS File**: `r params$gds_file`
- **Single-Cell Data File**: `r params$single_cell_file`
- **Counts Assay**: `r params$counts_assay`
- **Counts Slot**: `r params$counts_slot`
- **Cell-Type Column**: `r params$celltype_column`
- **Individual Column**: `r params$individual_column`
- **Minimum Cells for Pseudobulking**: `r params$min_cells`
- **Minimum Expression Percentage for Genes**: `r params$min_expression`
- **Cis Distance for eQTL Analysis**: `r params$cis_distance`
- **FDR Threshold for eQTL Results**: `r params$fdr_threshold`
- **Optimize Principal Components**: `r params$optimize_pcs`

### Total and Unique eGenes (5% FDR)

```{r plot_total_unique_eGenes, echo=FALSE}
total_genes <- function(cell_list) {
  sapply(cell_list, function(df) length(unique(df$gene)))
}

total_unique_genes <- function(cell_list) {
  sapply(cell_list, function(df) {
    # Genes in the current data frame
    current_genes <- unique(df$gene)
    
    # Genes in other data frames
    other_genes <- unique(unlist(sapply(cell_list, function(other_df) {
      if (identical(df, other_df)) return(NULL)
      return(other_df$gene)
    })))
    
    # Genes that are unique to the current data frame
    length(setdiff(current_genes, other_genes))
  })
}

mateqtlouts <- readRDS(params$eqtl_results)

mateqtlouts_filt <- lapply(mateqtlouts, function(x) {
  x <- x %>% 
    mutate(se = beta / t.stat, fstat = t.stat^2) 
  top_snps <- x %>%
    filter(FDR < 0.05) %>%
    group_by(gene) %>%
    arrange(p.value) %>%
    slice_head(n = 1) %>%
    ungroup() %>%
    arrange(FDR)
  return(top_snps)
})

# Get total eGenes (aggregate)
tot_eGenes <- total_genes(mateqtlouts_filt)
tot_unique_eGenes <- total_unique_genes(mateqtlouts_filt)
df <- data.frame(tot_eGenes, tot_unique_eGenes)
df$celltype <- rownames(df)
df$celltype <- gsub("_", " ", df$celltype)

# Get unique eGenes 
color_pal <- scales::hue_pal()(length(unique(df$celltype)))
colorvec <- setNames(color_pal, unique(df$celltype))

df$colorvec <- colorvec
tot_eGenes_sum <- lapply(mateqtlouts_filt, function(x) {
  return(x$gene)
})
tot_eGenes_sum <- length(unique(unlist(tot_eGenes_sum)))

df <- df[order(df$tot_eGenes, decreasing = TRUE), ]
df_long <- df %>%
  pivot_longer(cols = c(tot_eGenes, tot_unique_eGenes), names_to = "gene_type", values_to = "count")
df_long$gene_type <- factor(df_long$gene_type, levels = c("tot_unique_eGenes", "tot_eGenes"))
df_long$celltype_col <- paste0(df_long$celltype, df_long$gene_type)
df_long$celltype_col <- rep(c(2, 1), times = nrow(df_long) / 2)
df_long$celltype_col <- factor(df_long$celltype_col)
df_long$celltype <- factor(df_long$celltype, levels = rev(df$celltype))
df_long$celltype_col <- paste0(df_long$celltype, df_long$celltype_col)

colorvec <- df_long$colorvec
names(colorvec) <- df_long$celltype_col

make_lighter <- function(hex, fraction = 0.4) {
  rgb <- col2rgb(hex)
  new_rgb <- rgb + fraction * (255 - rgb)
  return(rgb(new_rgb[1], new_rgb[2], new_rgb[3], maxColorValue = 255))
}

colorvec <- df_long$colorvec
new_colors <- c(rbind(colorvec[c(TRUE, FALSE)], sapply(colorvec[c(TRUE, FALSE)], make_lighter)))
names(new_colors) <- df_long$celltype_col
max_val <- max(df_long$count) + 0.12 * max(df_long$count) 

g <- ggplot(df_long, aes(x = celltype, y = count, fill = celltype_col)) +
  geom_bar(stat = 'identity', position = position_dodge(), width = 0.8, colour = "black") +
  geom_text(aes(x = celltype, y = count, label = count), position = position_dodge(width = 0.9), hjust = -0.2, size = 6 / .pt) +
  scale_fill_manual(values = new_colors) +
  labs(y = "Total vs unique eGenes (5% FDR)") +
  coord_flip(ylim = c(NA, max_val)) +
  theme_classic() +
  theme(axis.text.x = element_text(size = 6.5, colour = "black"),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size = 6.5, colour = "black"),
        axis.title.x = element_text(size = 7, face = "bold"),
        legend.key.size = unit(0.7, "cm"),
        legend.position = "none",
        legend.text = element_text(size = 7))

print(g)
```

### Summary of eQTL Results

```{r summarize_eqtl_results, echo=FALSE}
eqtl_results_filtered <- readRDS(params$eqtl_results_filtered)

# Number of cell-types tested
num_celltypes <- length(unique(eqtl_results_filtered$celltype))
cat("Number of cell-types tested:", num_celltypes, "\n")

# Number of eGenes discovered (FDR < 0.05)
num_eGenes <- length(unique(eqtl_results_filtered$gene))
cat("Number of eGenes discovered (FDR < 0.05):", num_eGenes, "\n")

# Number of eGenes per cell-type
eGenes_per_celltype <- sapply(mateqtlouts_filt, function(df) length(unique(df$gene)))
cat("Number of eGenes per cell-type:\n")
print(eGenes_per_celltype)

# Number of unique eGenes per cell-type
unique_eGenes_per_celltype <- sapply(mateqtlouts_filt, function(df) {
  current_genes <- unique(df$gene)
  other_genes <- unique(unlist(sapply(mateqtlouts_filt, function(other_df) {
    if (identical(df, other_df)) return(NULL)
    return(other_df$gene)
  })))
  length(setdiff(current_genes, other_genes))
})
cat("Number of unique eGenes per cell-type:\n")
print(unique_eGenes_per_celltype)

# Number of shared eGenes
shared_eGenes <- num_eGenes - sum(unique_eGenes_per_celltype)
cat("Number of shared eGenes:", shared_eGenes, "\n")

# Cell-types sharing the most genes
shared_genes_matrix <- sapply(mateqtlouts_filt, function(df1) {
  sapply(mateqtlouts_filt, function(df2) {
    length(intersect(df1$gene, df2$gene))
  })
})
diag(shared_genes_matrix) <- 0
max_shared_genes <- max(shared_genes_matrix)
celltypes_sharing_most_genes <- which(shared_genes_matrix == max_shared_genes, arr.ind = TRUE)
cat("Cell-types sharing the most genes:", rownames(shared_genes_matrix)[celltypes_sharing_most_genes[1, 1]], "and", colnames(shared_genes_matrix)[celltypes_sharing_most_genes[1, 2]], "\n")
```

### Detailed eQTL Results

#### Filtered eQTL Results (FDR <= 0.05)

```{r filtered_eqtl_details, echo=FALSE}
cat("Below are the filtered eQTL results (FDR <= 0.05) for each unique gene in each cell type.\n")
for (celltype in names(mateqtlouts_filt)) {
  cat("\n### Cell Type:", celltype, "\n")
  unique_genes <- unique(mateqtlouts_filt[[celltype]]$gene)
  for (gene in unique_genes) {
    cat("\n#### Gene:", gene, "\n")
    print(mateqtlouts_filt[[celltype]] %>% filter(gene == gene))
  }
}
```

#### All eQTL Results

```{r all_eqtl_details, echo=FALSE}
cat("Below are the eQTL results for each unique gene in each cell type.\n")
for (celltype in names(mateqtlouts)) {
  cat("\n### Cell Type:", celltype, "\n")
  unique_genes <- unique(mateqtlouts[[celltype]]$gene)
  for (gene in unique_genes) {
    cat("\n#### Gene:", gene, "\n")
    print(mateqtlouts[[celltype]] %>% filter(gene == gene))
  }
}
```
