---
title: "Unified eQTL & PC Optimization Report"
output: html_document
params:
  eqtl_results_filtered: NULL
  eqtl_results: NULL
  genes_tested: NULL
  covs_used: NULL
  coarse_summaries: NULL
  fine_summaries: NULL
  outdir: NULL
  gds_file: NULL
  single_cell_file: NULL
  single_cell_file_list: NULL
  counts_assay: NULL
  counts_slot: NULL
  celltype_column: NULL
  individual_column: NULL
  min_cells: NULL
  min_expression: NULL
  cis_distance: NULL
  filter_chr: NULL
  fdr_threshold: NULL
  cov_file: NULL
  covariates_to_include: NULL
  subset_column: NULL
  subset_values: NULL
  optimize_pcs: NULL
  fixed_pcs: NULL
  pc_coarse_step: NULL
  pc_fine_step: NULL
  pc_fine_window: NULL
  pc_elbow_tol: NULL
  pc_early_stop_tol: NULL
  pc_early_stop_patience: NULL
  data_type: NULL
  workflow: NULL
  profile: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(data.table)
library(dplyr)
library(ggplot2)
library(tidyr)
library(purrr)
library(gridExtra)
library(scales)
library(DT)
format_cis_distance <- function(distance) {
  if (distance >= 1e6) {
    return(paste0(distance / 1e6, "Mb"))
  } else if (distance >= 1e3) {
    return(paste0(distance / 1e3, "Kb"))
  } else {
    return(paste0(distance, "bp"))
  }
}
data_type <- if (!is.null(params$data_type) && nzchar(params$data_type)) toupper(params$data_type) else "RNA"
analysis_descriptor <- if (data_type == "ATAC") {
  "This report summarizes ATAC-QTL analysis on pseudobulk peak counts, including peak-level normalization and QC."
} else {
  "This report summarizes eQTL analysis on pseudobulked RNA expression, including expression normalization and QC."
}
# Set theme for PC optimization plots
theme_set(theme_minimal() + theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

## eQTL Analysis Report

**Report Generated:** `r format(Sys.time(), "%d %B %Y, %H:%M:%S")`

This report was generated using the Nextflow eQTL Pipeline.

`r analysis_descriptor`

**GitHub Repository:** [https://github.com/johnsonlab-ic/sceQTL-flow](https://github.com/johnsonlab-ic/sceQTL-flow)

**Contact:** For questions or support regarding this pipeline, please contact Alexander Haglund (GitHub: [HaglundA](https://github.com/HaglundA))

---

### Pipeline Settings

The following settings were used in the pipeline:

```{r params-table, echo=FALSE}
params_tbl <- tibble::tibble(
  Parameter = c(
    "Output directory",
    "Workflow",
    "Profile",
    "GDS file",
    "Single-cell file",
    "Single-cell file list",
    "Counts assay",
    "Counts slot",
    "Cell-type column",
    "Individual column",
    "Min cells",
    "Min expression",
    "Cis distance",
    "Chromosomes",
    "FDR threshold",
    "Covariate file",
    "Covariates to include",
    "Subset column",
    "Subset values",
    "Optimize PCs",
    "Fixed PCs",
    "PC coarse step",
    "PC fine step",
    "PC fine window",
    "PC elbow tolerance",
    "PC early-stop tolerance",
    "PC early-stop patience"
  ),
  Value = c(
    params$outdir,
    params$workflow,
    params$profile,
    params$gds_file,
    params$single_cell_file,
    params$single_cell_file_list,
    params$counts_assay,
    params$counts_slot,
    params$celltype_column,
    params$individual_column,
    params$min_cells,
    params$min_expression,
    format_cis_distance(params$cis_distance),
    params$filter_chr,
    params$fdr_threshold,
    params$cov_file,
    params$covariates_to_include,
    params$subset_column,
    params$subset_values,
    params$optimize_pcs,
    params$fixed_pcs,
    params$pc_coarse_step,
    params$pc_fine_step,
    params$pc_fine_window,
    params$pc_elbow_tol,
    params$pc_early_stop_tol,
    params$pc_early_stop_patience
  )
)

knitr::kable(params_tbl, col.names = c("Parameter", "Value"))
```

```{r eqtl-load, echo=FALSE}
# Load the filtered eQTL results
if (!is.null(params$eqtl_results_filtered) && file.exists(params$eqtl_results_filtered)) {
  eqtl_results_filtered <- readRDS(params$eqtl_results_filtered)
} else {
  eqtl_results_filtered <- NULL
}

# Load the eQTL results
if (!is.null(params$eqtl_results) && file.exists(params$eqtl_results)) {
  eqtl_results <- readRDS(params$eqtl_results)
} else {
  eqtl_results <- NULL
}

# Load genes tested summary
if (!is.null(params$genes_tested) && file.exists(params$genes_tested)) {
  genes_tested_df <- data.table::fread(params$genes_tested, data.table = FALSE)
} else {
  genes_tested_df <- NULL
}

# Number and names of cell-types tested
cell_types <- if (!is.null(eqtl_results_filtered)) names(eqtl_results_filtered) else character(0)
num_celltypes <- length(cell_types)

celltype_colors <- if (length(cell_types) > 0) {
  setNames(scales::hue_pal()(length(cell_types)), cell_types)
} else {
  NULL
}

if (!is.null(eqtl_results) && num_celltypes > 0) {
  genes_list <- lapply(eqtl_results, function(df) unique(df$gene))
  all_genes <- unlist(genes_list)
  avg_genes_tested <- length(all_genes) / num_celltypes

  snps_list <- lapply(eqtl_results, function(df) unique(df$SNP))
  all_snps <- unique(unlist(snps_list))
  num_snps_tested <- length(all_snps)
} else {
  avg_genes_tested <- NA
  num_snps_tested <- NA
}

avg_genes_tested_display <- if (is.na(avg_genes_tested)) {
  NA
} else {
  formatC(round(avg_genes_tested), format = "f", digits = 0, big.mark = ",")
}

covs_used_files <- params$covs_used
covs_used_files <- covs_used_files[!is.null(covs_used_files) & covs_used_files != ""]

if (length(covs_used_files) > 0) {
  individuals_df <- map_df(covs_used_files, ~{
    if (file.exists(.x)) {
      celltype <- gsub("_covs_used.rds", "", basename(.x))
      covmat <- readRDS(.x)
      data.frame(celltype = celltype, n_individuals = ncol(covmat))
    } else {
      NULL
    }
  })
  if (nrow(individuals_df) == 0) {
    individuals_df <- NULL
  }
} else {
  individuals_df <- NULL
}

avg_individuals <- if (is.null(individuals_df)) NA else mean(individuals_df$n_individuals)
avg_individuals_display <- if (is.na(avg_individuals)) {
  NA
} else {
  formatC(round(avg_individuals), format = "f", digits = 0, big.mark = ",")
}
```

### Input Overview

- **Number of cell-types tested:** `r num_celltypes`
- **Names of cell-types tested:** `r paste(cell_types, collapse=", ")`
- **Average number of genes tested per cell-type:** `r avg_genes_tested_display`
- **Average number of individuals tested per cell-type:** `r avg_individuals_display`
- **Number of SNPs tested:** `r num_snps_tested`

```{r individuals-bar, fig.width=9, fig.height=4.5}
if (!is.null(individuals_df)) {
  plot_colors <- if (!is.null(celltype_colors)) {
    celltype_colors
  } else {
    setNames(scales::hue_pal()(length(unique(individuals_df$celltype))), unique(individuals_df$celltype))
  }
  individuals_df %>%
    mutate(celltype = factor(celltype, levels = names(plot_colors))) %>%
    ggplot(aes(x = celltype, y = n_individuals, fill = celltype)) +
    geom_col(width = 0.8, color = "black") +
    geom_text(aes(label = n_individuals), vjust = -0.4, size = 3) +
    scale_fill_manual(values = plot_colors, guide = "none") +
    labs(x = "Cell Type", y = "Number of Individuals") +
    theme_minimal() +
    theme(
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1),
      axis.title = element_text(size = 10)
    )
} else {
  cat("No covariate usage files available to summarize individuals.\n")
}
```

---

## Total and Unique eGenes (5% FDR)

```{r plot-total-unique-egenes, echo=FALSE}
if (!is.null(eqtl_results_filtered)) {
  total_genes <- function(cell_list) {
    sapply(cell_list, function(df) length(unique(df$gene)))
  }

  total_unique_genes <- function(cell_list) {
    sapply(cell_list, function(df) {
      current_genes <- unique(df$gene)
      other_genes <- unique(unlist(lapply(cell_list, function(other_df) {
        if (identical(df, other_df)) return(NULL)
        other_df$gene
      })))
      length(setdiff(current_genes, other_genes))
    })
  }

  mateqtlouts <- eqtl_results_filtered

  mateqtlouts_filt <- lapply(mateqtlouts, function(x) {
    x <- x %>% mutate(se = beta / t.stat, fstat = t.stat^2)
    x %>%
      group_by(gene) %>%
      arrange(p.value) %>%
      slice_head(n = 1) %>%
      ungroup() %>%
      arrange(FDR)
  })

  tot_eGenes <- total_genes(mateqtlouts_filt)
  tot_unique_eGenes <- total_unique_genes(mateqtlouts_filt)
  df <- data.frame(tot_eGenes, tot_unique_eGenes)
  df$celltype <- rownames(df)
  df$celltype <- gsub("_", " ", df$celltype)

  color_pal <- scales::hue_pal()(length(unique(df$celltype)))
  colorvec <- setNames(color_pal, unique(df$celltype))
  df$colorvec <- colorvec

  df <- df[order(df$tot_eGenes, decreasing = TRUE), ]
  df_long <- df %>%
    pivot_longer(cols = c(tot_eGenes, tot_unique_eGenes), names_to = "gene_type", values_to = "count")
  df_long$gene_type <- factor(df_long$gene_type, levels = c("tot_unique_eGenes", "tot_eGenes"))
  df_long$celltype_col <- rep(c(2, 1), times = nrow(df_long) / 2)
  df_long$celltype_col <- factor(df_long$celltype_col)
  df_long$celltype <- factor(df_long$celltype, levels = rev(df$celltype))
  df_long$celltype_col <- paste0(df_long$celltype, df_long$celltype_col)

  make_lighter <- function(hex, fraction = 0.4) {
    rgb <- col2rgb(hex)
    new_rgb <- rgb + fraction * (255 - rgb)
    rgb(new_rgb[1], new_rgb[2], new_rgb[3], maxColorValue = 255)
  }

  colorvec <- df_long$colorvec
  new_colors <- c(rbind(colorvec[c(TRUE, FALSE)], sapply(colorvec[c(TRUE, FALSE)], make_lighter)))
  names(new_colors) <- df_long$celltype_col
  max_val <- max(df_long$count) + 0.12 * max(df_long$count)

  g <- ggplot(df_long, aes(x = celltype, y = count, fill = celltype_col)) +
    geom_bar(stat = "identity", position = position_dodge(), width = 0.8, colour = "black") +
    geom_text(aes(label = count), position = position_dodge(width = 0.9), hjust = -0.2, size = 6 / .pt) +
    scale_fill_manual(values = new_colors) +
    labs(y = "Total vs unique eGenes (5% FDR)") +
    coord_flip(ylim = c(NA, max_val)) +
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 6.5, colour = "black"),
      axis.title.y = element_blank(),
      axis.text.y = element_text(size = 6.5, colour = "black"),
      axis.title.x = element_text(size = 7, face = "bold"),
      legend.position = "none"
    )

  print(g)
} else {
  cat("Filtered eQTL results not available for eGene plot.\n")
}
```

### total eGenes vs total genes tested

```{r plot-egenes-vs-tested, echo=FALSE, fig.width=10, fig.height=5}
if (!is.null(eqtl_results_filtered) && !is.null(genes_tested_df) && nrow(genes_tested_df) > 0) {
  egenes_discovered_df <- tibble(
    celltype = names(eqtl_results_filtered),
    egenes_discovered = vapply(eqtl_results_filtered, function(df) dplyr::n_distinct(df$gene), numeric(1))
  )

  plot_df <- egenes_discovered_df %>%
    inner_join(genes_tested_df, by = "celltype") %>%
    mutate(celltype = gsub("_", " ", celltype)) %>%
    arrange(desc(genes_tested))

  plot_long <- plot_df %>%
    pivot_longer(
      cols = c(egenes_discovered, genes_tested),
      names_to = "metric",
      values_to = "n_genes"
    ) %>%
    mutate(
      metric = dplyr::recode(metric,
        egenes_discovered = "eGenes discovered",
        genes_tested = "Total genes tested"
      ),
      celltype = factor(celltype, levels = unique(plot_df$celltype))
    )

  ggplot(plot_long, aes(x = celltype, y = n_genes, group = metric, color = metric)) +
    geom_line(size = 0.8) +
    geom_point(size = 2) +
    labs(x = "Cell Type", y = "Number of genes", color = "Metric") +
    theme_minimal() +
    theme(
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1),
      axis.title = element_text(size = 10)
    )
} else {
  cat("Genes tested summary is unavailable; cannot plot eGenes discovered vs total genes tested.\n")
}
```

### eQTL Summary Tables (Filtered at Configured FDR)

```{r eqtl-summary-tables, echo=FALSE}
if (!is.null(eqtl_results) && length(eqtl_results) > 0) {
  fdr_cutoff <- as.numeric(params$fdr_threshold)
  if (is.na(fdr_cutoff)) {
    fdr_cutoff <- 0.05
  }

  eqtl_filtered_dynamic <- lapply(eqtl_results, function(df) {
    if (!"FDR" %in% colnames(df)) return(df[0, , drop = FALSE])
    df %>% filter(FDR <= fdr_cutoff)
  })

  summary_tbl <- tibble(
    cell_type = names(eqtl_filtered_dynamic),
    total_unique_eGenes = vapply(eqtl_filtered_dynamic, function(df) dplyr::n_distinct(df$gene), numeric(1))
  ) %>%
    mutate(cell_type = gsub("_", " ", cell_type)) %>%
    arrange(desc(total_unique_eGenes))

  knitr::kable(
    summary_tbl,
    col.names = c("Cell Type", "Total Unique eGenes"),
    caption = paste0("Unique eGenes per cell type at FDR <= ", fdr_cutoff)
  )

  full_eqtl_dt <- purrr::imap_dfr(eqtl_filtered_dynamic, function(df, ct) {
    if (nrow(df) == 0) {
      return(tibble(
        cell_type = character(0),
        SNP = character(0),
        gene = character(0),
        effect_allele = character(0),
        other_allele = character(0),
        p.value = character(0),
        FDR = character(0),
        beta = character(0),
        se = character(0)
      ))
    }

    has_tstat <- "t.stat" %in% colnames(df)
    has_beta <- "beta" %in% colnames(df)
    se_vals <- rep(NA_real_, nrow(df))
    if (has_tstat && has_beta) {
      se_vals <- ifelse(df$t.stat == 0, NA_real_, df$beta / df$t.stat)
    }

    p_numeric <- as.numeric(df$p.value)
    fdr_numeric <- as.numeric(df$FDR)
    beta_numeric <- as.numeric(df$beta)
    gene_values <- as.character(df$gene)
    snp_values <- as.character(df$SNP)
    effect_values <- if ("effect_allele" %in% colnames(df)) as.character(df$effect_allele) else rep(NA_character_, nrow(df))
    other_values <- if ("other_allele" %in% colnames(df)) as.character(df$other_allele) else rep(NA_character_, nrow(df))

    top_per_gene <- tibble(
      gene = gene_values,
      SNP = snp_values,
      effect_allele = effect_values,
      other_allele = other_values,
      p_numeric = p_numeric,
      fdr_numeric = fdr_numeric,
      beta_numeric = beta_numeric,
      se_numeric = as.numeric(se_vals)
    ) %>%
      group_by(gene) %>%
      arrange(p_numeric, .by_group = TRUE) %>%
      slice_head(n = 1) %>%
      ungroup()

    tibble(
      cell_type = gsub("_", " ", ct),
      SNP = top_per_gene$SNP,
      gene = top_per_gene$gene,
      effect_allele = top_per_gene$effect_allele,
      other_allele = top_per_gene$other_allele,
      p.value = formatC(top_per_gene$p_numeric, format = "e", digits = 2),
      FDR = formatC(top_per_gene$fdr_numeric, format = "e", digits = 2),
      beta = as.character(signif(top_per_gene$beta_numeric, 3)),
      se = as.character(signif(top_per_gene$se_numeric, 3))
    )
  })

  if (nrow(full_eqtl_dt) > 0) {
    DT::datatable(
      full_eqtl_dt,
      rownames = FALSE,
      filter = "top",
      extensions = "Buttons",
      options = list(
        pageLength = 20,
        lengthMenu = c(10, 25, 50, 100),
        dom = "Bfrtip",
        buttons = c("copy", "csv")
      ),
      caption = htmltools::tags$caption(
        style = "caption-side: top; text-align: left;",
        paste0("Unique-gene eQTL table by cell type (best association per gene, FDR <= ", fdr_cutoff, ")")
      )
    )
  } else {
    cat("No eQTL associations passed the configured FDR threshold for DT summary.\n")
  }
} else {
  cat("eQTL results are unavailable; cannot build summary tables.\n")
}
```

# PC Optimization Results

```{r pcopt-setup, echo=FALSE}
# Use the passed coarse and fine summary files instead of searching in outdir
coarse_summary_files <- params$coarse_summaries
fine_summary_files <- params$fine_summaries

# Filter out any empty or NULL values
coarse_summary_files <- coarse_summary_files[!is.null(coarse_summary_files) & coarse_summary_files != ""]
fine_summary_files <- fine_summary_files[!is.null(fine_summary_files) & fine_summary_files != ""]

cat(sprintf("Received %d coarse summary files\n", length(coarse_summary_files)))
cat(sprintf("Received %d fine summary files\n", length(fine_summary_files)))
```

## PC Optimization Strategy

This section details the two-stage PC optimization process used in the eQTL pipeline.

### Overview

The pipeline employs a **coarse-to-fine grid search** with **early stopping** and **elbow tolerance** to select the optimal number of principal components (PCs) for each cell type:

1. **Coarse Stage**: Test PCs in steps of 10 (2, 12, 22, ...) up to 50% of sample size
2. **Early Stopping**: Halt coarse evaluation when 2 consecutive steps show <1% improvement
3. **Fine Stage**: Refine search around the best coarse PC (±10 in steps of 2)
4. **Elbow Selection**: Pick the smallest PC count within 2% of maximum associations

---

## Coarse Grid Optimization

```{r coarse-load}
# Load coarse summary files passed from the workflow
if (length(coarse_summary_files) > 0 && !is.null(coarse_summary_files)) {
  coarse_data <- map_df(coarse_summary_files, ~{
    if (file.exists(.x)) {
      celltype <- gsub("_coarse_summary.csv", "", basename(.x))
      fread(.x) %>% mutate(celltype = celltype)
    } else {
      NULL
    }
  }, .id = "file") %>%
    select(-file) %>%
    as_tibble()
  
  if (nrow(coarse_data) == 0) {
    coarse_data <- NULL
  }
} else {
  cat("No coarse summary files provided from workflow.\n")
  coarse_data <- NULL
}
```

The coarse grid evaluates a broad range of PC values in large steps (default step size = 10) to quickly identify the region with the best eQTL power. Early stopping halts the search when two consecutive steps show less than 1% improvement in associations, preventing unnecessary computation. The plots below show the total number of associations and percent gains at each PC value tested, with the best PC (selected for fine-tuning) highlighted in red.

### Coarse Grid: Total Associations

```{r coarse-assoc-plot, fig.width=14, fig.height=10}
if (!is.null(coarse_data)) {
  plot_colors <- if (exists("celltype_colors")) {
    celltype_colors
  } else {
    setNames(scales::hue_pal()(length(unique(coarse_data$celltype))), unique(coarse_data$celltype))
  }
  coarse_data %>%
    mutate(celltype = factor(celltype, levels = names(plot_colors))) %>%
    arrange(n_pcs) %>%
    ggplot(aes(x = n_pcs, y = n_assoc, color = celltype)) +
    geom_line(size = 0.6) +
    geom_point(size = 2) +
    geom_point(data = coarse_data %>% filter(is_best), size = 2.8, color = "#d62728") +
    scale_color_manual(values = plot_colors, guide = "none") +
    facet_wrap(~celltype, scales = "free") +
    labs(x = "Number of PCs", y = "Number of Associations") +
    theme_minimal() +
    theme(
      panel.grid.minor = element_blank(),
      strip.text = element_text(size = 10, face = "bold"),
      axis.text = element_text(size = 9),
      axis.title = element_text(size = 10)
    )
} else {
  cat("No coarse data for visualization.\n")
}
```

### Coarse Grid: Percent Gains

```{r coarse-gains-plot, fig.width=14, fig.height=10}
if (!is.null(coarse_data)) {
  plot_colors <- if (exists("celltype_colors")) {
    celltype_colors
  } else {
    setNames(scales::hue_pal()(length(unique(coarse_data$celltype))), unique(coarse_data$celltype))
  }
  coarse_data %>%
    mutate(celltype = factor(celltype, levels = names(plot_colors))) %>%
    arrange(n_pcs) %>%
    ggplot(aes(x = n_pcs, y = gain_pct, color = celltype)) +
    geom_line(size = 0.6) +
    geom_point(size = 2) +
    geom_point(data = coarse_data %>% filter(is_best), size = 2.8, color = "#d62728") +
    scale_color_manual(values = plot_colors, guide = "none") +
    facet_wrap(~celltype, scales = "free") +
    labs(x = "Number of PCs", y = "Gain (%)") +
    theme_minimal() +
    theme(
      panel.grid.minor = element_blank(),
      strip.text = element_text(size = 10, face = "bold"),
      axis.text = element_text(size = 9),
      axis.title = element_text(size = 10)
    )
} else {
  cat("No coarse data for visualization.\n")
}
```

---

## Fine Grid Optimization

```{r fine-load}
# Load fine summary files passed from the workflow
if (length(fine_summary_files) > 0 && !is.null(fine_summary_files)) {
  fine_data <- map_df(fine_summary_files, ~{
    if (file.exists(.x)) {
      celltype <- gsub("_fine_summary.csv", "", basename(.x))
      fread(.x) %>% mutate(celltype = celltype)
    } else {
      NULL
    }
  }, .id = "file") %>%
    select(-file) %>%
    as_tibble()
  
  if (nrow(fine_data) == 0) {
    fine_data <- NULL
  }
} else {
  cat("No fine summary files provided from workflow.\n")
  fine_data <- NULL
}
```

## Fine Grid Optimization

The fine grid refines the search around the best coarse PC, testing in smaller steps (default step size = 2) within a defined window (default ±10 PCs). The elbow rule then selects the smallest PC count that falls within 2% of the maximum observed associations, balancing eQTL power with model parsimony. The plot below shows the number of associations at each fine-grid PC, with the selected PC highlighted in red.

### Fine Grid Visualization with Elbow Rule

```{r fine-plot, fig.width=14, fig.height=10}
if (!is.null(fine_data)) {
  plot_colors <- if (exists("celltype_colors")) {
    celltype_colors
  } else {
    setNames(scales::hue_pal()(length(unique(fine_data$celltype))), unique(fine_data$celltype))
  }
  fine_data %>%
    mutate(celltype = factor(celltype, levels = names(plot_colors))) %>%
    arrange(n_pcs) %>%
    ggplot(aes(x = n_pcs, y = n_assoc, color = celltype)) +
    geom_line(size = 0.6) +
    geom_point(size = 2) +
    geom_point(data = fine_data %>% filter(is_selected == TRUE), 
               size = 3.5, color = "#d62728", shape = 17) +
    scale_color_manual(values = plot_colors, guide = "none") +
    facet_wrap(~celltype, scales = "free") +
    labs(x = "Number of PCs", y = "Number of eGenes") +
    theme_minimal() +
    theme(
      panel.grid.minor = element_blank(),
      strip.text = element_text(size = 10, face = "bold"),
      axis.text = element_text(size = 9),
      axis.title = element_text(size = 10)
    )
} else {
  cat("No fine data for visualization.\n")
}
```



## Parameters Used

```{r pc-params-table}
params_summary <- tibble(
  Parameter = c("Coarse Step", "Fine Step", "Fine Window", "Early Stop Tolerance", "Early Stop Patience", "Elbow Tolerance"),
  Value = c("10", "2", "±10", "1%", "2 consecutive steps", "2%"),
  Description = c(
    "PC increment for coarse grid",
    "PC increment for fine grid",
    "Range around best coarse PC",
    "Threshold for low gain detection",
    "Consecutive low-gain steps to trigger stop",
    "Allow up to 2% below maximum associations"
  )
)
knitr::kable(params_summary, caption = "PC Optimization Parameters")
```

```{r final-pcs-by-celltype, echo=FALSE}
if (!is.null(fine_data) && nrow(fine_data) > 0 && "is_selected" %in% colnames(fine_data)) {
  final_pcs_tbl <- fine_data %>%
    filter(is_selected == TRUE) %>%
    select(celltype, n_pcs) %>%
    distinct() %>%
    mutate(celltype = gsub("_", " ", celltype)) %>%
    arrange(celltype)
} else if (!is.null(cell_types) && length(cell_types) > 0 && !is.null(params$fixed_pcs)) {
  final_pcs_tbl <- tibble(
    celltype = gsub("_", " ", cell_types),
    n_pcs = as.integer(params$fixed_pcs)
  ) %>%
    arrange(celltype)
} else {
  final_pcs_tbl <- NULL
}

if (!is.null(final_pcs_tbl) && nrow(final_pcs_tbl) > 0) {
  knitr::kable(
    final_pcs_tbl,
    col.names = c("Cell Type", "Final PCs Used"),
    caption = "Final number of PCs used per cell type"
  )
} else {
  cat("Final PC selection table is not available.\n")
}
```

---

## Interpretation Guide

### Coarse Stage
- **Early Stopping**: If 2 consecutive steps show <1% improvement in associations, coarse evaluation halts
- **Ceiling**: The last PC value before early stopping triggers
- **Best Coarse PC**: The PC with maximum associations **up to the ceiling**

### Fine Stage
- **Elbow Tolerance**: Region within 2% of the maximum associations
- **Selected PC**: The smallest PC count in the elbow region (balances accuracy with parsimony)
- **Interpretation**: This selection avoids overfitting (too many PCs) while capturing nearly-optimal eQTL power

### Example
If coarse shows gains of 10%, 4%, 0.8%, 0.2%:
- Early stop triggers at PC showing 0.2% (2 consecutive <1%)
- Ceiling = PC at 0.2%
- Best coarse = PC with highest associations before ceiling
- Fine grid = ±10 around best coarse
- Final = smallest PC in fine grid within 2% of max

---

# eQTL Results Summary

... (rest of the original final_report.Rmd content, e.g., eGene plots, tables, etc.) ...
