---
title: "sceQTL-flow — cis-eQTL Report"
subtitle: "MatrixEQTL · per cell type"
date: now
date-format: "DD MMMM YYYY, HH:mm"
format:
  html:
    theme: flatly
    toc: true
    toc-location: left
    toc-depth: 3
    toc-title: "Contents"
    code-fold: true
    code-tools: true
    embed-resources: true
    fig-width: 9
    fig-height: 5
    fig-dpi: 110
    tbl-cap-location: top
engine: knitr
execute:
  echo: false
  warning: false
  message: false
params:
  eqtl_results_filtered: null
  eqtl_summary: null
  cells_per_individual: null
  outdir: null
  gds_file: null
  single_cell_file: null
  single_cell_file_list: null
  counts_assay: null
  counts_slot: null
  celltype_column: null
  individual_column: null
  min_cells: null
  min_expression: null
  cis_distance: null
  filter_chr: null
  fdr_threshold: null
  cov_file: null
  covariates_to_include: null
  subset_column: null
  subset_values: null
  optimize_pcs: null
  fixed_pcs: null
  pc_coarse_step: null
  pc_fine_step: null
  pc_fine_window: null
  pc_elbow_tol: null
  pc_early_stop_tol: null
  pc_early_stop_patience: null
  standardize_residuals: null
  workflow: null
  profile: null
---

```{r setup}
library(data.table)
library(dplyr)
library(ggplot2)
library(tidyr)
library(purrr)
library(knitr)

# ggsci NPG palette (nrc) — hardcoded so the report needs no ggsci dependency
npg_pal <- c("#E64B35", "#4DBBD5", "#00A087", "#3C5488", "#F39B7F",
             "#8491B4", "#91D1C2", "#DC0000", "#7E6148", "#B09C85")

celltype_palette <- function(cts) {
  cts <- sort(unique(as.character(cts)))
  setNames(rep(npg_pal, length.out = length(cts)), cts)
}

theme_report <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title       = element_text(face = "bold"),
      strip.text       = element_text(face = "bold"),
      axis.title       = element_text(face = "bold")
    )
}
theme_set(theme_report())

fmt_int <- function(x) formatC(round(as.numeric(x)), format = "f", digits = 0, big.mark = ",")

format_cis_distance <- function(d) {
  d <- as.numeric(d)
  if (is.na(d)) return(NA_character_)
  if (d >= 1e6) paste0(d / 1e6, "Mb") else if (d >= 1e3) paste0(d / 1e3, "Kb") else paste0(d, "bp")
}

is_set <- function(x) !is.null(x) && !(tolower(as.character(x)) %in% c("none", "", "na"))
```

```{r load-data}
eqtl_results_filtered <- if (is_set(params$eqtl_results_filtered) && file.exists(params$eqtl_results_filtered)) {
  readRDS(params$eqtl_results_filtered)
} else NULL

eqtl_summary <- if (is_set(params$eqtl_summary) && file.exists(params$eqtl_summary)) {
  as.data.frame(readRDS(params$eqtl_summary))
} else NULL

cells_per_individual <- if (is_set(params$cells_per_individual) && file.exists(params$cells_per_individual)) {
  fread(params$cells_per_individual)
} else NULL

cell_types     <- if (!is.null(eqtl_results_filtered)) names(eqtl_results_filtered) else
                  if (!is.null(eqtl_summary)) eqtl_summary$celltype else character(0)
num_celltypes  <- length(cell_types)
ct_colors      <- if (num_celltypes > 0) celltype_palette(cell_types) else NULL

optimize_pcs <- isTRUE(as.logical(params$optimize_pcs))
```

## Overview

```{r overview-values}
n_ind_range <- if (!is.null(eqtl_summary)) {
  r <- range(eqtl_summary$n_individuals)
  if (r[1] == r[2]) as.character(r[1]) else paste0(r[1], "–", r[2])
} else NA
tot_egenes  <- if (!is.null(eqtl_summary)) fmt_int(sum(eqtl_summary$n_egenes)) else NA
tot_pairs   <- if (!is.null(eqtl_summary)) fmt_int(sum(eqtl_summary$n_sig_pairs)) else NA
snps_tested <- if (!is.null(eqtl_summary)) fmt_int(max(eqtl_summary$n_snps_tested)) else NA
```

::: {.callout-note icon=false}
## Run at a glance

- **Cell types tested:** `r num_celltypes` — `r paste(cell_types, collapse = ", ")`
- **Individuals per cell type:** `r n_ind_range`
- **SNPs tested:** `r snps_tested`
- **Significant eGenes (≤ `r params$fdr_threshold` FDR, summed across cell types):** `r tot_egenes`
- **Significant SNP–gene pairs:** `r tot_pairs`
- **PC strategy:** `r if (optimize_pcs) "optimized (coarse→fine grid search)" else paste0("fixed (", params$fixed_pcs, " PCs)")`
:::

This report was generated by [sceQTL-flow](https://github.com/johnsonlab-ic/sceQTL-flow). Contact: Alexander Haglund ([HaglundA](https://github.com/HaglundA)).

## Pipeline settings

```{r params-table}
# Resolve the single-cell input to a single, accurate description (the old
# report always printed both "Single-cell file" and "...file list" rows).
list_used <- is_set(params$single_cell_file_list)
sc_paths  <- if (list_used) trimws(strsplit(params$single_cell_file_list, ",")[[1]]) else params$single_cell_file
sc_fmt <- {
  ext <- tolower(tools::file_ext(sc_paths[1]))
  switch(ext, h5ad = "AnnData (.h5ad)", rds = "Seurat (.rds)", ext)
}
sc_desc <- if (list_used) {
  sprintf("%s — %d files", sc_fmt, length(sc_paths))
} else {
  sprintf("%s — %s", sc_fmt, basename(as.character(sc_paths)))
}

rows <- list(
  c("Workflow",              params$workflow),
  c("Profile",              params$profile),
  c("Output directory",     params$outdir),
  c("GDS file",             basename(as.character(params$gds_file))),
  c("Single-cell input",    sc_desc),
  c("Counts assay / slot",  paste0(params$counts_assay, " / ", params$counts_slot)),
  c("Cell-type column",     params$celltype_column),
  c("Individual column",    params$individual_column),
  c("Min cells (pseudobulk)", params$min_cells),
  c("Min expression",       params$min_expression),
  c("Cis distance",         format_cis_distance(params$cis_distance)),
  c("Chromosomes",          params$filter_chr),
  c("FDR threshold",        params$fdr_threshold)
)
if (is_set(params$cov_file)) {
  rows <- c(rows, list(
    c("Covariate file",        basename(as.character(params$cov_file))),
    c("Covariates included",   params$covariates_to_include)))
}
if (is_set(params$subset_column)) {
  rows <- c(rows, list(
    c("Subset column",  params$subset_column),
    c("Subset values",  params$subset_values)))
}
rows <- c(rows, list(
  c("PC selection", if (optimize_pcs) "optimized" else paste0("fixed (", params$fixed_pcs, ")")),
  c("Standardize final residuals (SdY=1)", if (isTRUE(as.logical(params$standardize_residuals))) "YES" else "NO")))

params_tbl <- as.data.frame(do.call(rbind, rows), stringsAsFactors = FALSE)
colnames(params_tbl) <- c("Parameter", "Value")
kable(params_tbl)
```

## Statistical corrections

The expression matrix passes through the following corrections, in this order, before an association
is tested at each SNP-gene pair. The diagram shows the pipeline's full canonical shape; the table below
states which of these steps were actually applied for this run.

```{r corrections-flags}
standardize_residuals <- isTRUE(as.logical(params$standardize_residuals))
has_cov <- is_set(params$cov_file)
```

```{mermaid}
%%| fig-cap: "Corrections applied to the expression matrix before eQTL testing"
flowchart TD
    A[Raw pseudobulk counts] --> B["Normalize: log2(CPM + 1)"]
    B --> C["Regress out covariates (per gene lm)"]
    C --> D["Compute hidden-factor PCs<br/>(prcomp, scale.=TRUE)"]
    D --> E["Regress out PCs (per gene lm)"]
    E --> F["Center + scale to unit variance<br/>(SdY = 1)"]
    F --> G[MatrixEQTL cis association test]
```

The canonical pipeline always has this shape; the table below shows which steps were actually applied for this run.

```{r corrections-table}
n_pcs_used <- {
  cov_files <- list.files("covs", pattern = "_covs_used\\.rds$", full.names = TRUE)
  if (length(cov_files) == 0) {
    NA_integer_
  } else {
    vals <- vapply(cov_files, function(f) {
      m <- readRDS(f)
      if (is.null(m)) 0L else nrow(m)
    }, integer(1))
    vals
  }
}
pcs_used_any <- length(n_pcs_used) > 0 && !all(is.na(n_pcs_used)) && any(n_pcs_used > 0, na.rm = TRUE)

corrections_rows <- list(
  c("1. Normalize (log2(CPM+1))", "Always applied"),
  c("2. Regress out covariates",
    if (has_cov) paste0("Applied — ", params$covariates_to_include) else "Skipped — no --cov_file given"),
  c("3. Compute hidden-factor PCs",
    if (pcs_used_any) paste0("Applied — ", if (optimize_pcs) "optimized per cell type" else paste0("fixed (", params$fixed_pcs, ")")) else "Skipped — 0 PCs"),
  c("4. Regress out PCs", if (pcs_used_any) "Applied" else "Skipped — no PCs selected"),
  c("5. Center + scale to unit variance (SdY=1)",
    if (standardize_residuals) "Applied — final residuals are mean-centered, unit-variance per gene" else "Skipped — residuals left on log2(CPM+1) scale (--standardize_residuals false)")
)
corrections_tbl <- as.data.frame(do.call(rbind, corrections_rows), stringsAsFactors = FALSE)
colnames(corrections_tbl) <- c("Step", "This run")
kable(corrections_tbl)
```

```{r sdy-check}
sdy_files <- list.files("sdy", pattern = "_sdY_check\\.csv$", full.names = TRUE)
if (length(sdy_files) > 0) {
  sdy_df <- rbindlist(lapply(sdy_files, function(f) {
    ct <- gsub("_sdY_check\\.csv$", "", basename(f))
    d <- fread(f)
    data.table(celltype = ct, median_sd = median(d$sd, na.rm = TRUE),
               p10 = quantile(d$sd, 0.1, na.rm = TRUE), p90 = quantile(d$sd, 0.9, na.rm = TRUE))
  }))
  cap <- if (standardize_residuals) {
    "Achieved per-gene residual SD, by cell type (median and 10th/90th percentile). Should sit at 1.0 — this is what makes SdY=1 valid to pass directly to coloc."
  } else {
    "Achieved per-gene residual SD, by cell type (median and 10th/90th percentile). Standardization is off for this run, so these are NOT 1 — the true SD must be estimated separately before using SdY=1 in coloc."
  }
  kable(sdy_df, digits = 3, caption = cap)
} else {
  cat("No SdY check files available.\n")
}
```

## Sample sizes

```{r individuals}
ind_df <- if (!is.null(eqtl_summary)) {
  eqtl_summary %>% select(celltype, n_individuals)
} else NULL

if (!is.null(ind_df)) {
  ind_df %>%
    mutate(celltype = factor(celltype, levels = celltype[order(n_individuals, decreasing = TRUE)])) %>%
    ggplot(aes(celltype, n_individuals, fill = celltype)) +
    geom_col(width = 0.8, colour = "black") +
    geom_text(aes(label = n_individuals), vjust = -0.4, size = 3.3) +
    scale_fill_manual(values = ct_colors, guide = "none") +
    labs(x = NULL, y = "Individuals") +
    theme(axis.text.x = element_text(angle = 35, hjust = 1))
} else {
  cat("No eQTL summary available.\n")
}
```

### Cells per individual

```{r cells-per-individual, fig.height=6}
if (!is.null(cells_per_individual) && nrow(cells_per_individual) > 0) {
  below_lbl <- sprintf("Below min_cells (%s)", params$min_cells)
  dropped_lbl <- "Cell type dropped (<15 individuals)"
  alpha_vals <- setNames(c(1, 0.4, 0.15), c("Included", below_lbl, dropped_lbl))

  cpi_colors <- celltype_palette(cells_per_individual$celltype)

  cpi <- cells_per_individual %>%
    mutate(
      status = case_when(
        celltype_dropped   ~ dropped_lbl,
        !kept_min_cells    ~ below_lbl,
        TRUE                ~ "Included"
      ),
      status = factor(status, levels = c("Included", below_lbl, dropped_lbl))
    ) %>%
    group_by(celltype) %>%
    mutate(individual = factor(individual, levels = individual[order(n_cells, decreasing = TRUE)])) %>%
    ungroup()

  ggplot(cpi, aes(individual, n_cells, fill = celltype, alpha = status)) +
    geom_col(width = 0.85, colour = "black", linewidth = 0.15) +
    geom_hline(yintercept = as.numeric(params$min_cells), linetype = "dashed", colour = "grey30") +
    facet_wrap(~celltype, scales = "free_x") +
    scale_fill_manual(values = cpi_colors, guide = "none") +
    scale_alpha_manual(values = alpha_vals, name = NULL) +
    scale_y_log10() +
    labs(x = "Individuals (sorted by cell count)", y = "Cells per individual (log scale)") +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
          legend.position = "top")
} else {
  cat("No per-individual cell-count data available.\n")
}
```

```{r cells-per-individual-table}
if (!is.null(cells_per_individual) && nrow(cells_per_individual) > 0) {
  cpi_tab <- cells_per_individual %>%
    group_by(celltype) %>%
    summarise(
      `Individuals seen`   = n(),
      `Below min_cells`    = sum(!kept_min_cells),
      `Median cells`       = fmt_int(median(n_cells)),
      `Min cells`          = fmt_int(min(n_cells)),
      `Max cells`          = fmt_int(max(n_cells)),
      `Cell type dropped`  = any(celltype_dropped),
      .groups = "drop"
    ) %>%
    arrange(`Cell type dropped`, desc(`Individuals seen`))
  kable(cpi_tab, align = c("l", "r", "r", "r", "r", "r", "c"),
        caption = sprintf("Cell-count QC per cell type. min_cells=%s; cell types with <15 individuals after that filter are dropped from eQTL mapping entirely.", params$min_cells))
} else {
  cat("No per-individual cell-count data available.\n")
}
```

## eQTL results

```{r results-table}
if (!is.null(eqtl_summary)) {
  tab <- eqtl_summary %>%
    arrange(desc(n_egenes)) %>%
    transmute(
      `Cell type`      = celltype,
      Individuals      = n_individuals,
      `Genes tested`   = fmt_int(n_genes_tested),
      `Sig. pairs`     = fmt_int(n_sig_pairs),
      eGenes           = fmt_int(n_egenes)
    )
  kable(tab, align = c("l", "r", "r", "r", "r"),
        caption = sprintf("Significant cis-eQTLs per cell type (≤ %s FDR).", params$fdr_threshold))
} else {
  cat("No eQTL summary available.\n")
}
```

### Total and cell-type-unique eGenes

```{r egene-plot, fig.height=5.5}
if (!is.null(eqtl_results_filtered)) {
  # one row per gene (top SNP), then total vs cell-type-private eGenes
  top_by_gene <- lapply(eqtl_results_filtered, function(x) {
    x %>% group_by(gene) %>% slice_min(p.value, n = 1, with_ties = FALSE) %>% ungroup()
  })
  gene_sets <- lapply(top_by_gene, function(x) unique(x$gene))

  unique_counts <- sapply(names(gene_sets), function(ct) {
    others <- unlist(gene_sets[setdiff(names(gene_sets), ct)])
    length(setdiff(gene_sets[[ct]], others))
  })

  egene_df <- tibble(
    celltype = names(gene_sets),
    Total    = sapply(gene_sets, length),
    Unique   = unique_counts
  ) %>%
    arrange(desc(Total)) %>%
    mutate(celltype = factor(celltype, levels = rev(celltype))) %>%
    pivot_longer(c(Total, Unique), names_to = "type", values_to = "count") %>%
    mutate(type = factor(type, levels = c("Total", "Unique")))

  ggplot(egene_df, aes(celltype, count, fill = celltype, alpha = type)) +
    geom_col(position = position_dodge(width = 0.8), width = 0.75, colour = "black") +
    geom_text(aes(label = count), position = position_dodge(width = 0.8),
              hjust = -0.15, size = 2.9) +
    scale_fill_manual(values = ct_colors, guide = "none") +
    scale_alpha_manual(values = c(Total = 1, Unique = 0.45), name = NULL) +
    coord_flip(clip = "off") +
    expand_limits(y = max(egene_df$count) * 1.12) +
    labs(x = NULL, y = "eGenes (≤ FDR threshold)") +
    theme(legend.position = "top")
} else {
  cat("Filtered eQTL results not available.\n")
}
```

`Total` = all eGenes for the cell type; `Unique` = eGenes significant in that cell type only.

```{r child-pcopt, child = if (optimize_pcs) "pc_optimization.qmd" else NULL}
```
